### 2022-09-06 doc/src/CSRs.tex

\begin{commentary}
The addresses for all CSRs in this document have been carefully
selected not to conflict with existing allocations, or with other
likely extensions, so far as is known.
\end{commentary}

@@

For RV32, the \emph{high-half} CSRs listed in the table extend
registers \z{mideleg}, \z{mie}, \z{mvien}, \z{mvip}, and \z{mip} to
support a total of 64 interrupt causes.
Each low-half register (e.g. \z{mideleg}) covers interrupt numbers
0--31, one bit per interrupt number, while its high-half partner
(\z{midelegh}) covers interrupt numbers 32--63, again one bit per
interrupt number.
For RV64, a single \mbox{64-bit} CSR (\z{mideleg}) covers the same set
of interrupt numbers, 0--63.
The Advanced Interrupt Architecture requires that these high-half CSRs
exist for RV32, but they may all be merely read-only zeros.

>>>>

@@

For RV32, the \emph{high-half} CSRs listed in the table extend
registers \z{mideleg}, \z{mie}, \z{mvien}, \z{mvip}, and \z{mip} to
support a total of 64 interrupt causes.
Each low-half register (e.g.\ \z{mideleg}) covers interrupt numbers
0--31, one bit per interrupt number, while its high-half partner
(\z{midelegh}) covers interrupt numbers 32--63, again one bit per
interrupt number.
For RV64, a single \mbox{64-bit} CSR (\z{mideleg}) covers the same set
of interrupt numbers, 0--63.
The Advanced Interrupt Architecture requires that these high-half CSRs
exist for RV32, but they may all be merely read-only zeros.

### 2022-09-06 doc/src/IMSIC.tex

For each privilege level and each virtual hart to which MSIs
may be directed at a hart, the hart's IMSIC contains a separate
\emph{interrupt file}.
Assuming a hart implements supervisor mode, its IMSIC has at least two
interrupt files, one for machine level and the other for supervisor
level.
When a hart also implements the hypervisor extension, its IMSIC may
have additional interrupt files for virtual harts.
The number of interrupt files an IMSIC has for virtual harts is exactly
\emph{GEILEN}, the number of supported guest external interrupts.

@@

\begin{commentary}
Platform standards may increase the minimum number of interrupt
identities that must be implemented by each interrupt file.
Platform standards for UNIX-like operating systems are likely
to require a minimum of either 127 or 255 distinct identities at
supervisor level.
\end{commentary}

@@

A platform might provide a means for software to configure the sizes
of interrupt files in some way, such as by allowing a smaller size at
machine level to be traded for a larger size at supervisor level, or
vice versa.
Any such configurability is outside the scope of this specification.
It is recommended, however, that only machine level be given the power
to change the sizes of interrupt files.

@@

\begin{commentary}
An interrupt file's lowest identity numbers have been given the
highest priorities, not the reverse order, because it is only for
the highest-priority interrupts that priority order may need to be
carefully managed, yet it is the low-numbered identities, 1~through 63
(or perhaps 1 through 255), that are guaranteed to exist across all
systems.
Consider, for example, that an interrupt file's highest-priority
interrupt---presumably the most time-critical---is always identity
number~1.
If priority order were reversed, the highest-priority interrupt would
have different identity numbers on different machines, depending on how
many identities are implemented by interrupt files.
The ability for software to assign fixed identity numbers to the
highest-priority interrupts was considered worth any discomfort that
may be felt from interrupt priorities being the reverse of the natural
number order.
\end{commentary}

@@

If an IMSIC contains a supervisor-level interrupt file and software at
the attached hart enables \mbox{S-mode} that was previously disabled
(e.g. by changing bit~S of CSR \z{misa} from zero to one), all state
of the supervisor-level interrupt file is valid and consistent but
otherwise {\unspecified}.
Likewise, if an IMSIC contains guest interrupt files and software at
the attached hart enables the hypervisor extension that was previously
disabled (e.g. by changing bit~H of \z{misa} from zero to one), all
state of the IMSIC's guest interrupt files is valid and consistent but
otherwise {\unspecified}.

@@

For machine level, the relevant CSRs are:\nopagebreak
\begin{displayLinesTable}
\z{miselect} \qquad \z{mireg} \qquad \z{mtopei} \\
\end{displayLinesTable}

When supervisor mode is implemented, the set of supervisor-level CSRs
matches those of machine level:
\begin{displayLinesTable}
\z{siselect} \qquad \z{sireg} \qquad \z{stopei} \\
\end{displayLinesTable}

And when the hypervisor extension is implemented, there is a
corresponding set of VS CSRs:
\begin{displayLinesTable}
\z{vsiselect} \qquad \z{vsireg} \qquad \z{vstopei} \\
\end{displayLinesTable}

@@

The combined read and write of \z{mtopei} or \z{stopei} in the second
step can be done by a single CSRRW machine instruction,
\begin{displayLinesTable}
\z{csrrw }\textit{rd}\z{,} \z{mtopei} or \z{stopei,} \z{x0} \\
\end{displayLinesTable}
where \textit{rd} is the destination register for value~\z{i}.

>>>>

For each privilege level and each virtual hart to which MSIs
may be directed at a hart, the hart's IMSIC contains a separate
\emph{interrupt file}.
Assuming a hart implements supervisor mode, its IMSIC has at least two
interrupt files, one for machine level and the other for supervisor
level.
When a hart also implements the hypervisor extension, its IMSIC may
have additional interrupt files for virtual harts,
called \emph{guest interrupt files}.
The number of guest interrupt files an IMSIC has for virtual harts is exactly
\emph{GEILEN}, the number of supported guest external interrupts,
as defined for the hypervisor extension
by the {\RISCV} Privileged Architecture.

@@

\begin{commentary}
Platform standards may increase the minimum number of interrupt
identities that must be implemented by each interrupt file.
\end{commentary}

@@

A platform might provide a means for software to configure the
number of interrupt files in an IMSIC and/or their sizes, such as
by allowing a smaller interrupt file at machine level to be traded
for a larger one at supervisor level, or vice versa, for example.
Any such configurability is outside the scope of this specification.
It is recommended, however, that only machine level be given the power
to change the number and sizes of interrupt files in an IMSIC.

@@

\begin{commentary}
An interrupt file's lowest identity numbers have been given the
highest priorities, not the reverse order, because it is only for
the highest-priority interrupts that priority order may need to be
carefully managed, yet it is the low-numbered identities, 1~through 63
(or perhaps 1 through 255), that are guaranteed to exist across all
systems.
Consider, for example, that an interrupt file's highest-priority
interrupt---presumably the most time-critical---is always identity
number~1.
If priority order were reversed, the highest-priority interrupt would
have different identity numbers on different machines, depending on how
many identities are implemented by interrupt files.
The ability for software to assign fixed identity numbers to the
highest-priority interrupts is considered worth any discomfort that
may be felt from interrupt priorities being the reverse of the natural
number order.
\end{commentary}

@@

If an IMSIC contains a supervisor-level interrupt file and software at
the attached hart enables \mbox{S-mode} that was previously disabled
(e.g.\ by changing bit~S of CSR \z{misa} from zero to one), all state
of the supervisor-level interrupt file is valid and consistent but
otherwise {\unspecified}.
Likewise, if an IMSIC contains guest interrupt files and software at
the attached hart enables the hypervisor extension that was previously
disabled (e.g.\ by changing bit~H of \z{misa} from zero to one), all
state of the IMSIC's guest interrupt files is valid and consistent but
otherwise {\unspecified}.

@@

For machine level, the relevant CSRs are
\z{miselect}, \z{mireg}, and \z{mtopei}.
When supervisor mode is implemented, the set of supervisor-level CSRs
matches those of machine level:
\z{siselect}, \z{sireg}, and \z{stopei}.
And when the hypervisor extension is implemented,
there are three corresponding VS CSRs:
\z{vsiselect}, \z{vsireg}, and \z{vstopei}.

@@

The combined read and write of \z{mtopei} or \z{stopei} in the second
step can be done by a single CSRRW machine instruction,
\begin{displayLinesTable}
\z{csrrw }\textit{rd}\z{,} \z{mtopei}/\z{stopei,} \z{x0} \\
\end{displayLinesTable}
where \textit{rd} is the destination register for value~\z{i}.

### 2022-09-06 doc/src/IOMMU.tex

\item
At the IMSIC interrupt file, save to memory the existing values of
registers \z{eidelivery} and \z{eithreshold}, and set \z{eidelivery}
=~0.

\item
Modify all relevant translation tables at \mbox{IOMMU}s so that MSIs for
this virtual interrupt file are now stored in the MRIF.
If necessary, synchronize with all \mbox{IOMMU}s to ensure that no straggler
MSIs will arrive at the old IMSIC interrupt file after this step.

@@

\item
At the IMSIC interrupt file, load registers \z{eithreshold} and
\z{eidelivery} with the values that were earlier saved.

@@

First, if the I/O devices that a virtual hart controls are behind
multiple \mbox{IOMMUs}, then multiple MRIF structures are needed, one per
\mbox{IOMMU}, not just a single MRIF structure.
Furthermore, in addition to locations for storing \z{eidelivery} and
\z{eithreshold}, software needs a place for a complete copy of the
interrupt file's implemented \z{eip} array, apart from the MRIFs.
While a virtual interrupt file is in memory, its interrupt-pending bits
will be split across all the MRIFs and the saved \z{eip} array.
The interrupt-enable bits will exist only in the MRIFs.

@@

\item
At the IMSIC interrupt file, save to memory the existing values of
registers \z{eidelivery} and \z{eithreshold}, and set \z{eidelivery}
=~0.

\item
At each \mbox{IOMMU}, modify all relevant translation tables so that MSIs
for this virtual interrupt file are now stored in the individual MRIF
matched to the \mbox{IOMMU}.
If necessary, synchronize with all \mbox{IOMMU}s to ensure that no straggler
MSIs will arrive at the old IMSIC interrupt file after this step.

@@

When a write is identified as an MSI from a device, an
\emph{interrupt file number} is extracted from the original guest
physical address as
\begin{displayLinesTable}
interrupt file number = extract($A$\,\z{>>}\,12, address mask) \\
\end{displayLinesTable}
The extract function here is the same generic bit extract performed
by {\RISCV} instruction BEXT, defined by the ``Bitmanip'' extension
(B~extension).
The bit extract function extract($x$,~$y$) discards all bits from
$x$ whose matching bits in the same positions in the mask~$y$ are
zeros, and packs the remaining bits from $x$ contiguously at the
least-significant end of the result, keeping the same bit order as $x$
and filling any other bits at the most-significant end of the result
with zeros.
For example, if the bits of $x$ and $y$ are
\begin{displayLinesTable}[r@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c]
$x$ = & a & b & c & d & e & f & g & h \\
$y$ = & 1 & 0 & 1 & 0 & 0 & 1 & 1 & 0 \\
\end{displayLinesTable}
then the value of extract($x$,~$y$) has bits 0~0~0~0~a~c~f~g.

@@

The number of entries in an MSI page table is always a power of two,
specifically $\mbox{2}^{k}$ where $k$ is the number of bits that are
ones in the MSI address mask used to extract the interrupt file number
from the destination guest physical address.
If an MSI page table has 256 or fewer entries, the start of the table
is always aligned to a \mbox{4-KiB} page address in real physical
memory.
If an MSI page table has ${\mbox{2}^{k} > \mbox{256}}$
entries, the table must be naturally aligned to a
$\mbox{2}^{k}\times \mbox{16-byte}$ address boundary.
If an MSI page table is not aligned as required, all entries in the
table appear to an \mbox{IOMMU} as {\unspecified}, and any address an
\mbox{IOMMU} may compute and use for reading an individual MSI PTE from the
table is also {\unspecified}.

>>>>

\item
Save to memory the existing values of the IMSIC interrupt file's
registers \z{eidelivery} and \z{eithreshold}, and set \z{eidelivery}
=~0.

\item
Modify all relevant translation tables at \mbox{IOMMU}s so that MSIs for
this virtual interrupt file are now stored in the MRIF.
If necessary, synchronize with all \mbox{IOMMU}s to ensure that no straggler
MSIs will arrive at the IMSIC interrupt file after this step.

@@

\item
Load the IMSIC interrupt file's registers \z{eithreshold} and
\z{eidelivery} with the values that were earlier saved.

@@

First, if the I/O devices that a virtual hart controls are behind
multiple \mbox{IOMMUs}, then multiple MRIF structures are needed, one per
\mbox{IOMMU}, not just a single MRIF structure.
Furthermore, in addition to locations for storing \z{eidelivery} and
\z{eithreshold}, software needs a place for a complete copy of the
interrupt file's implemented \z{eip} array, apart from the MRIFs.
While a virtual interrupt file is in memory, its interrupt-pending bits
will be split across all the MRIFs and the saved \z{eip} array.
The interrupt-enable bits may exist only in the MRIFs.

@@

\item
Save to memory the existing values of the IMSIC interrupt file's
registers \z{eidelivery} and \z{eithreshold}, and set \z{eidelivery}
=~0.

\item
At each \mbox{IOMMU}, modify all relevant translation tables so that MSIs
for this virtual interrupt file are now stored in the individual MRIF
matched to the \mbox{IOMMU}.
If necessary, synchronize with all \mbox{IOMMU}s to ensure that no straggler
MSIs will arrive at the IMSIC interrupt file after this step.

@@

When a write is identified as an MSI from a device, an
\emph{interrupt file number} is extracted from the original guest
physical address as
\begin{displayLinesTable}
interrupt file number = extract($A$\,\z{>>}\,12, address mask) \\
\end{displayLinesTable}
Here, extract($x$,~$y$) is a ``bit extract'' that discards all bits from
$x$ whose matching bits in the same positions in the mask~$y$ are
zeros, and packs the remaining bits from $x$ contiguously at the
least-significant end of the result, keeping the same bit order as $x$
and filling any other bits at the most-significant end of the result
with zeros.
For example, if the bits of $x$ and $y$ are
\begin{displayLinesTable}[r@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c]
$x$ = & a & b & c & d & e & f & g & h \\
$y$ = & 1 & 0 & 1 & 0 & 0 & 1 & 1 & 0 \\
\end{displayLinesTable}
then the value of extract($x$,~$y$) has bits 0~0~0~0~a~c~f~g.

@@

The number of entries in an MSI page table is
$\mbox{2}^{k}$ where $k$ is the number of bits that are
ones in the MSI address mask used to extract the interrupt file number
from the destination guest physical address.
If an MSI page table has 256 or fewer entries, the start of the table
is aligned to a \mbox{4-KiB} page address in real physical
memory.
If an MSI page table has ${\mbox{2}^{k} > \mbox{256}}$
entries, the table must be naturally aligned to a
$\mbox{2}^{k}\times \mbox{16-byte}$ address boundary.
If an MSI page table is not aligned as required, all entries in the
table appear to an \mbox{IOMMU} as {\unspecified}, and any address an
\mbox{IOMMU} may compute and use for reading an individual MSI PTE from the
table is also {\unspecified}.

### 2022-09-06 doc/src/IPIs.tex

The same kind of mechanism (but with a different set of addresses) may
or may not exist for signaling supervisor-level software interrupts
(major code~1) at remote harts as well.
If not directly supported in this way, then a supervisor-level software
interrupt is typically sent to another hart through an environment call
from supervisor mode to machine mode.
An operating system running in \mbox{S-mode} thus invokes a specific
SBI function for delivering a software interrupt to another hart,
causing machine-level software at the originating hart to send a
machine-level IPI to the destination hart, where software then sets the
supervisor-level software interrupt-pending bit (SSIP) in CSR \z{mip}.

When harts have IMSICs, instead of using the Privileged Architecture's
mechanism for signaling software interrupts at remote harts, an IPI
can be sent to a hart by writing to the destination hart's IMSIC, the
same as a regular message-signaled interrupt (MSI)\@.
In that case, an incoming IPI appears at the destination hart as an
\emph{external interrupt} routed through the IMSIC, rather than as a
software interrupt as before.
However, so long as the same software (e.g.\@ an operating system or
machine monitor) is in control at both endpoints of an IPI, source
and destination, there should be no reason for a destination hart
to misinterpret the purpose of an incoming external interrupt that
represents an IPI.

If harts do not have IMSICs, then the method specified by the {\RISCV}
Privileged Architecture is assumed to be used for IPIs, with software
interrupts signaled at destination harts.
On the other hand, when harts have IMSICs, the machinery for triggering
software interrupts at remote harts is redundant with the capabilities
of the IMSICs, so it is downgraded from a requirement to an option,
useful perhaps only to maximize software compatibility across a range
of {\RISCV} systems, with and without IMSICs.
If a machine implements IMSICs and not the earlier software-interrupt
mechanism, then the bits of CSRs \z{mip} and \z{mie} for machine-level
software interrupts, MSIP and MSIE, are hardwired to zero in harts.

@@

Similarly, SBI functions for sending IPIs can easily continue to
be supported without clients being at all aware of a change in the
underlying hardware for delivering IPIs between harts.
\end{commentary}

>>>>

The same kind of mechanism (but with a different set of memory addresses) may
or may not exist for signaling supervisor-level software interrupts
(major code~1) at remote harts as well.
If not directly supported in this way, a supervisor-level software
interrupt is typically sent to another hart instead through an environment call
from supervisor mode to machine mode.
An operating system running in \mbox{S-mode} thus invokes a specific
SBI function for delivering a software interrupt to another hart,
causing machine-level software at the originating hart to send a
machine-level IPI to the destination hart, where software then sets the
supervisor-level software interrupt-pending bit (SSIP) in CSR \z{mip}.

When harts have IMSICs, instead of using the Privileged Architecture's
mechanism for signaling software interrupts at remote harts, an IPI
can be sent to a hart by writing to the destination hart's IMSIC, the
same as a regular message-signaled interrupt (MSI)\@.
In that case, an incoming IPI appears at the destination hart as an
\emph{external interrupt} routed through the IMSIC, rather than as a
software interrupt as before.
However, so long as the same software (e.g.\ an operating system or
machine monitor) is in control at both endpoints of an IPI, source
and destination, there should be no reason for a destination hart
to misinterpret the purpose of an incoming external interrupt that
represents an IPI.

If harts do not have IMSICs, then the method specified by the {\RISCV}
Privileged Architecture is assumed to be used for IPIs, signaling software
interrupts at destination harts.
On the other hand, when harts have IMSICs, the machinery for triggering
software interrupts at remote harts is redundant with the capabilities
of the IMSICs, so it is downgraded from a requirement to an option,
useful perhaps only to provide software compatibility across a range
of {\RISCV} systems, with and without IMSICs.
If a machine implements IMSICs and not the earlier software-interrupt
mechanism, then the bits of CSRs \z{mip} and \z{mie} for machine-level
software interrupts, MSIP and MSIE, are hardwired to zero in harts.

@@

Similarly, SBI environment calls for sending IPIs can easily continue to
be supported without clients being at all aware of a change in the
underlying hardware for delivering IPIs between harts.
\end{commentary}

### 2022-09-06 doc/src/VSLevel.tex
Add preface and update status of each chapter

\item
if \z{hvictl}.VTI =~0:
\begin{tightList}
\item[+]
all pending-and-enabled major interrupts indicated by
\z{vsip}/\z{vsiph} and \z{vsie}/\z{vsieh} other than a supervisor
external interrupt (code 9), with priority numbers assigned by
\z{hviprio1}/\z{hviprio1h} and \z{hviprio2}/\z{hviprio2h};
\end{tightList}

>>>>

\item
if \z{hvictl}.VTI =~0:
\begin{tightList}
\item[+]
the highest-priority pending-and-enabled major interrupt indicated by
\z{vsip}/\z{vsiph} and \z{vsie}/\z{vsieh} other than a supervisor
external interrupt (code 9), using the priority numbers assigned by
\z{hviprio1}/\z{hviprio1h} and \z{hviprio2}/\z{hviprio2h};
\end{tightList}

### 2022-09-06 doc/src/intro.tex

In addition to external interrupts from I/O devices, the {\RISCV}
Privileged Architecture specifies a few other major classes of
interrupts for harts.
The Privileged Architecture's timer interrupts remain supported
in full, and software interrupts remain at least partly supported,
although neither appears in Figures \ref{fig:intrsWithoutIMSICs}
and~\ref{fig:intrsWithIMSICs}.
For the specifics on software interrupts, refer to
Chapter~\ref{ch:IPIs}, ``Interprocessor interrupts (IPIs).''

@@

\begin{commentary}
We contend that the need to communicate an I/O interrupt to multiple
harts is sufficiently rare that standard hardware support for multicast
cannot be justified in this case.
\end{commentary}

@@

\begin{commentary}
The original Platform-Level Interrupt Controller (PLIC) for {\RISCV}
is configurable so each interrupt source signals external interrupts to
any subset of the harts, potentially all harts.
When multiple harts receive an external interrupt from a single cause
at the PLIC, the first hart to \emph{claim} the interrupt at the PLIC
is the one responsible for servicing it.
Usually this sets up a race, where the subset of harts configured to
receive the multicast interrupt all take an external interrupt trap
simultaneously and compete to be the first to claim the interrupt at
the PLIC.
The intention is to provide a form of \mbox{1-of-$N$} interrupt
delivery;
however, for all the harts that fail to win the claim, the interrupt
trap becomes wasted effort.

@@

As revealed in subsequent chapters, the exact set
of CSRs and behavior added by the AIA, and hence
implied by Smaia or Ssaia, depends on some other factors:
the base ISA's XLEN (RV32 or RV64), whether \mbox{S-mode}
and the hypervisor extension are implemented,
and whether the hart has an IMSIC.
But individual AIA extension names are not
provided for each possible valid subset.
Rather, the different combinations are inferable
from the intersection of features indicated (such as
RV64I + \mbox{S-mode} + Smaia, but without the hypervisor extension).

>>>>

In addition to external interrupts from I/O devices, the {\RISCV}
Privileged Architecture specifies a few other major classes of
interrupts for harts.
The Privileged Architecture's timer interrupts remain supported
in full, and software interrupts remain at least partly supported,
although neither appears in Figures \ref{fig:intrsWithoutIMSICs}
and~\ref{fig:intrsWithIMSICs}.
For the specifics on software interrupts, refer to
Chapter~\ref{ch:IPIs}, ``Interprocessor Interrupts (IPIs).''

@@

\begin{commentary}
We contend that the need to communicate an I/O interrupt to multiple
harts is sufficiently rare that standardizing hardware support for multicast
cannot be justified in this case.
\end{commentary}

@@

\begin{commentary}
The original Platform-Level Interrupt Controller (PLIC) for {\RISCV}
is configurable so each interrupt source signals external interrupts to
any subset of the harts, potentially all harts.
When multiple harts receive an external interrupt from a single cause
at the PLIC, the first hart to \emph{claim} the interrupt at the PLIC
is the one responsible for servicing it.
Usually this sets up a race, where the subset of harts configured to
receive the multicast interrupt all take an external interrupt trap
simultaneously and compete to be the first to claim the interrupt at
the PLIC.
The intention is to provide a form of \mbox{1-of-$N$} interrupt
delivery.
However, for all the harts that fail to win the claim, the interrupt
trap becomes wasted effort.

@@

As revealed in subsequent chapters, the exact set
of CSRs and behavior added by the AIA, and hence
implied by Smaia or Ssaia, depends on
the base ISA's XLEN (RV32 or RV64), on whether \mbox{S-mode}
and the hypervisor extension are implemented,
and on whether the hart has an IMSIC.
But individual AIA extension names are not
provided for each possible valid subset.
Rather, the different combinations are inferable
from the intersection of features indicated (such as
RV64I + \mbox{S-mode} + Smaia, but without the hypervisor extension).
